Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-12-08 paper/conclusion.tex
small changes -- still need to go over wording in discussion section

We examined the rotational evolution of \kepler\ field stars using their
velocity dispersions as a proxy for age.
We found that old groups of cool dwarfs selected to be coeval using the
\citet{angus2019} gyrochronology relation do {\it not} have the same velocity
dispersion across all temperatures.
This implies that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-color/\teff\ relation for old stars.
It appears that the period-color/\teff\ relation changes shape over time in a
way that qualitatively agrees with theoretical that which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.
At young ages, rotation period is anti-correlated with \teff: cooler stars
spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
We speculate that the rotation period gap \citep{mcquillan2014} may separate
a young regime where stellar rotation periods decrease with increasing mass
from an old regime where periods increase with increasing mass, however more
data are needed to provide a conclusive result.
% The period-color/\teff\ relation seems to start flattening out after $\sim$1
% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
% older than the period gap which is located at a gyrochronal age of around
% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
Finally, when velocity dispersion is interpreted as an age proxy, it appears
that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
suggests that lower-mass stars remain active for longer, allowing their
rotation periods to be measured at older ages.
% We outlined a number of scenarios which could provide alternative
% explanations for these observations, including incorrect dust corrections for
% the lowest-mass stars and an excess of companions increasing the velocity
% dispersion for these stars.

>>>>

We examined the rotational evolution of \kepler\ field stars by interpreting
their velocity dispersions as a proxy for age.
% We found that old groups of cool dwarfs, selected to be coeval using the
% \citet{angus2019} gyrochronology relation, do {\it not} have the same velocity
% dispersion across all temperatures.
We found that the \citet{angus2019} relation, which is based on the
period-color relation of the 650 Myr Praesepe cluster, does not correctly
describe the period-age-color/\teff\ relation for old stars.
It appears that the period-color/\teff\ relation changes shape over time in a
way that qualitatively agrees with theoretical models which include a
mass-dependent core-envelope angular momentum transport \citep{spada2019}.
At young ages, rotation period is anti-correlated with \teff: cooler stars
spin more slowly than hotter stars of the same age.
However, at intermediate ages the relation flattens out and K dwarfs rotate at
the same rate, regardless of mass.
At old ages, it seems that cooler K dwarfs spin more rapidly than hotter K
dwarfs of the same age.
% We speculate that the rotation period gap \citep{mcquillan2014} may separate
% a young regime where stellar rotation periods decrease with increasing mass
% from an old regime where periods increase with increasing mass, however more
% data are needed to provide a conclusive result.
% The period-color/\teff\ relation seems to start flattening out after $\sim$1
% Gyr (see figure \ref{fig:age_cut}), potentially around the same age, or just
% older than the period gap which is located at a gyrochronal age of around
% $\sim$1.1 Gyr (see figure \ref{fig:dispersion_period_teff}).
Finally, when velocity dispersion is interpreted as an age proxy, it appears
that the oldest stars in the \mct\ catalog are cooler than 4500 K, which
suggests that lower-mass stars remain active for longer, allowing their
rotation periods to be measured at older ages.
% We outlined a number of scenarios which could provide alternative
% explanations for these observations, including incorrect dust corrections for
% the lowest-mass stars and an excess of companions increasing the velocity
% dispersion for these stars.

### 2019-12-08 paper/results.tex
small changes -- still need to go over wording in discussion section

% word limit: 500
\section{Results}
\label{sec:results}

@@

Figure \ref{fig:masses_with_models} shows the same as figure \ref{fig:vplot},
however mass is plotted on the x-axis instead of temperature and the models
are those of \citet{spada2019}.
The masses of these stars are taken from the \kepler\ input catalog
\citep{KIC}.
The \citet{spada2019} models appear to qualitatively agree with the data:
their rotation period-\teff\ relation flattens out over time and eventually
inverts over a narrow range of temperatures.
The models reproduce the `dip' feature seen in the data.

@@

\begin{figure}
  \caption{
    Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
    x-axis.
    Masses are from the \kepler\ input catalog.
    The white lines show the \citet{spada2019} rotational evolution models at
    0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
    period.
    These models include age and mass-dependent coupling between the stellar
    core and envelope.
    The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
    an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
    decreases with decreasing mass.
    A similar phenomenon is visible in the velocity dispersions of field stars
    shown as colored points in the background of this figure.
}
  \centering
    \includegraphics[width=1\textwidth]{masses_with_models}
\label{fig:masses_with_models}
\end{figure}

>>>>

% word limit: 500
\section{Results and Discussion}
\label{sec:results}

@@

Figure \ref{fig:masses_with_models} shows the same as figure \ref{fig:vplot},
however mass is plotted on the x-axis instead of temperature and the models
are those of \citet{spada2019}.
The masses of these stars are taken from the \kepler\ input catalog
\citep{brown2011}.
The \citet{spada2019} models appear to qualitatively agree with the data:
their rotation period-\teff\ relation flattens out over time and eventually
inverts over a narrow range of temperatures.

These results indicate that stars of spectral type ranging from late G to late
K follow a braking law that changes over time.
In particular, the relationship between rotation period and effective
temperature appears to flatten out and eventually invert.
These results provide further evidence for `stalled' rotational evolution of K
dwarfs, like that observed in open clusters \citep{curtis2019} and reproduced
by models that vary angular momentum transport between stellar core and
envelope with time and mass \citep{spada2019}.

The velocity dispersions of stars in the \mct\ sample provide the following
picture of rotational evolution.
At young ages, stellar rotation period decreases with mass, likely because
lower-mass stars with deeper convection zones have stronger magnetic fields,
larger Alfv\'en radii and therefore experience greater angular momentum loss
rate.
According to the \citet{spada2019} model, there is minimal transportation of
angular momentum from the surface to the core of the star at these young ages,
so the surface slows down but the core keeps spinning rapidly.
At intermediate ages, rotation period is constant with mass, and at late ages
rotation period {\it increases} with mass for GK dwarfs.
The explanation for this, according to the \citet{spada2019} model, is that
lower-mass stars are still braking more efficiently at these intermediate and
old ages but their cores are more tightly coupled to their envelopes, allowing
angular momentum transport between the two interior layers.
Angular momentum resurfaces and prevents the stellar envelopes from
spinning-down rapidly.

@@

% \begin{figure}
%   \caption{
%     Similar to figure \ref{fig:vplot}, with mass instead of \teff\ on the
%     x-axis.
%     Masses are from the \kepler\ input catalog.
%     The white lines show the \citet{spada2019} rotational evolution models at
%     0.5, 1, 1.5, 2, 2.5, 4 and 4.57 Gyr, where age increases with rotation
%     period.
%     These models include age and mass-dependent coupling between the stellar
%     core and envelope.
%     The oldest models (4 and 4.57 Gyrs), at the largest rotation periods, show
%     an inversion at $\sim$0.7-0.5 M$_\odot$, where rotation period briefly
%     decreases with decreasing mass.
%     A similar phenomenon is visible in the velocity dispersions of field stars
%     shown as colored points in the background of this figure.
% }
%   \centering
%     \includegraphics[width=1\textwidth]{masses_with_models}
% \label{fig:masses_with_models}
% \end{figure}
